---
title: "Dimensionality reduction techniques"
author: Matthias Döring
date: '2018-11-13T15:00:00Z'
description: "TODO"
draft: true
categories:
  - machine-learning
tags:
    - dimensionality reduction
thumbnail: "/post/machine-learning/linear_models_cover.png"

---

Dimensionality reduction has two primary use cases: data exploration and machine learning. It is useful for data exploration because dimensionality reduction to few dimensions (e.g. 2 or 3 dimensions) allows for visualizing the samples. Such a visualization can then be used to obtain insights from the data (e.g. detect clusters and identify outliers). For machine learning, dimensionality reduction is relevant when there are many features, particularly when there are more features than samples. By reducing the size of the feature space, it is possible to obtain models that generalize better since the signal-to-noise ratio is improved. 

## Loading a whiskey data set

I have previously used a data set describing the characteristics of whiskeys to [draw radar plots](/post/data-visualization/radar-plot/). Here, we will the whiskey data set to identify how different dimensionality reduction techniques perform on this data set.

```{r, message = FALSE}
library(RCurl)
# load data as character
f <- getURL('https://www.datascienceblog.net/data-sets/whiskies.csv')
# read table from text connection
df <- read.csv(textConnection(f), header=T)
head(df)
# select characterics of the whiskeys
features <- c("Body", "Sweetness", "Smoky",
            "Medicinal", "Tobacco", "Honey",
            "Spicy", "Winey", "Nutty",
            "Malty", "Fruity", "Floral")
feat.df <- df[, c("Distillery", features)]
```

## Assumptions about the results

Before we begin reducing the dimensionality of the data, we should think about what kind of results we would like to obtain. For the whiskey data, we would like that whiskeys with similar properties are close to each other in the reduced space.

Distilleries that are in proximity to one another should produce whiskeys that exhibit some similaries. To validate this assumptions, we are going to test whether the mean whiskey characteristics differ between distilleries from different regions. So, let us run a MANOVA test:

```{r}
m <- manova(as.matrix(df[, features]) ~ Region, df)
summary(m)
```

The test statistic seems to be significant, so we can reject the null hypothesis (there is no effect of region on the characteristics).

## Geographical locations of the distilleries

Thus, if we plot the distilleries by ```Latitude``` and ```Longitude```, this should give us some idea of how a reasonable dimensionality reduction should look like. The scotch whiskey regions are the following:

![Scotch regions (Licensed under CC BY-SA 3.0 and retrieved from https://commons.wikimedia.org/wiki/File:Scotch_regions.svg)](https://upload.wikimedia.org/wikipedia/commons/f/fd/Scotch_regions.svg)

We will replicate this plot with our data set.


Let us see where the distilleries are located by plotting their latitude and longitude coordinates: 

```{r, message = FALSE, fig.width = 10, fig.height = 10}
library(ggplot2)
library(ggrepel) # for geom_text_repel: smart text labels
uk.map <- map_data ("world", region = "UK") 
scotland.map <- uk.map[uk.map$subregion == "Scotland",]
p <- ggplot(data = scotland.map, aes(x = long, y = lat)) + 
geom_map(map = uk.map, 
       aes(map_id = region),
       fill="white", colour = "black") +
coord_map() + 
geom_point(data = df, aes(y = long, x = lat, color = Region),
     alpha = .75) +
ggtitle ("Locations of Scottish Whiskey Distilleries")
# for storing the map with labels:
#geom_text_repel(data = map.df, aes(y = long, x = lat, label = Distillery)) 
# ggsave(file.path("distillery_map.png"), p, units = "cm", height = 80, width = 80)
p
```

I also created a [high-resolution version of the distillery map](/post/machine-learning/distillery_map.png) where the labels of the distilleries are annotated.

## PCA


### Results of PCA

Using PCA, I would assign three clusters to the data:

```{r, fig.width = 10, fig.height = 10}
library(cluster)
library(ggfortify)
data <- df[,features]
rownames(data) <- df$Distillery
cl <- pam(data, 3)
autoplot(cl, frame = TRUE, frame.type = 'norm', label = TRUE)
```

The principal components seem to reflect the following characteristics:

* PC1 seems to indicate a notion of a *heavy taste*: i.e. full body, smokiness, medicinal taste (e.g. Laphroaig, Ardbeg, or Lagavulin vs Auchentoshan or Abelour)
* PC2 seems to indicate a *balanced taste*: i.e. a well-rounded taste profile where no characteristic stands out (e.g. Glenfiddich or Auchentoshan vs Glendronach or Macallan)

Let us check whether the clusters overrepresent certain regions:

```{r}
tabs <- vector("list", 3)
for (i in seq(3)) {
    idx <- which(cl$clustering == i)
    regions <- df$Region[idx]
    tabs[[i]] <- table(regions)
}
print(tabs)
```

From these data, we can interpret the clustering in the following way:

* Cluster 1: Whiskeys with a more complex taste, mostly from the Highlands/Speyside
* Cluster 2: Well-balanced/mild Whiskeys, mostly from Speyside and Highlands
* Cluster 3: Smoky whiskeys, highly overrepresented for whiskeys from Islay

### Multi-Dimensional Scaling

This is basically the same as PCA, so cut it?

```{r}
autoplot(cmdscale(daisy(data), eig = TRUE), label = TRUE, label.size = 3)
```

## t-SNE

t-SNE can be performed in the following way:

```{r, message = FALSE}
library(Rtsne)
set.seed(1211) # fix seed for reproducibility of t-SNE
tsne <- Rtsne(data, dims = 2, perplexity = 5)
t.df <- as.data.frame(tsne$Y)
colnames(t.df) <- c("V1", "V2")
t.df <- cbind(t.df, Cluster = factor(cl$clustering))
t.df$Distillery <- rownames(t.df)
ggplot(t.df, aes(x = V1, y = V2, color = Cluster, label = Distillery)) +
    geom_point() + geom_text()
```

The result is a bit more clearer with t-SNE than with PCA and the two dimensions have slightly different interpretations than with PCA:

* V1 seems to indicate how balanced/intense the taste is. The outliers here are the smoky Islay whiskeys on the left (e.g. Lagavulin) as well as some of the richer Highland whiskeys on the right (e.g. Macallan). The whiskeys in the middle such as Glenfiddich have a very balanced taste profile.
* V2: I cannot figure that out.
