---
title: "Interpreting Generalized Linear Models"
author: Matthias Döring
date: '2018-11-08T15:00:00Z'
draft: true
description: "TODO"
categories:
  - machine-learning
tags:
thumbnail: "/post/machine-learning/linear_models_cover.png"

---

## Data preparation

We will take 70% of samples for training and 30% for testing:

```{r}
data(airquality)
ozone <- subset(na.omit(airquality), 
        select = c("Ozone", "Solar.R", "Wind", "Temp"))
set.seed(123)
N.train <- ceiling(0.7 * nrow(ozone))
N.test <- nrow(ozone) - N.train
trainset <- sample(seq_len(nrow(ozone)), N.train)
testset <- setdiff(seq_len(nrow(ozone)), trainset)
```

## Interpretation of generalized linear models

Interpreting generalized linear models (GLM) obtained through ```glm``` is similar to [interpreting conventional linear models](/post/machine-learning/linear_models/). Still, there are some differences. Let us first build a GLM that corresponds to our previous model:

```{r}
model.pois <- glm(Ozone ~ Solar.R + Temp + Wind, data = ozone, 
                family = "poisson", subset = trainset)
summary(model.pois)
```

# Residuals:
https://stackoverflow.com/questions/2531489/understanding-glmresiduals-and-residglm
http://people.stat.sfu.ca/~raltman/stat402/402L11.pdf
http://data.princeton.edu/wws509/notes/c3s8.html

# Null deviance:

https://www.youtube.com/watch?v=JC56jS2gVUE
https://stats.stackexchange.com/questions/108995/interpreting-residual-and-null-deviance-in-glm-r
