---
title: "Linear and Quadratic Discriminant Analysis"
author: Matthias Döring
date: '2018-11-29'
description: ""
draft: true
categories:
  - machine-learning
tags:
    - supervised-learning
    - linear-model
    - R
---
Discriminant analysis encompasses methods that can be used for both classification and dimensionality reduction. There are three flavors of discriminant analysis: linear discriminant analysis (LDA), quadratic discriminant analysis (QDA), and regularized discriminant analysis (RDA). The main driver for the popularity of discriminant analysis is that these methods implicitly transform the data into low-dimensional space.

## The approach of LDA

LDA can be explained through two perspectives. The first is the probabilistic, Gaussian interpretation and the second is the interpretation as a transformation that is due to Fisher. The first interpretation is useful to understand the underlying assumptions of LDA. The second interpretation gives a better inutition on how the inherent dimensionality reduction of LDA works.

### Probabilistic interpretation

Each class $k \in \{1, \ldots, K\}$ is assigned a prior $\hat{\pi}_k such that that $\sum_{i=1}^k \hat{\pi}_k = 1$. The posterior probability is

\[\rm{Pr}(G = k |X = x) = \frac{f_k(x) \pi_k}{\sum_{l=1}^K f_l(x) \pi_l} \]

where $f_k(x)$ is the density of $X$ conditioned on $k$. The maximum-a-posterior estimator is

\[G(x) = \arg \max_k \rm{Pr}(G = k | X = x) = \arg \max_k f_k(x) \pi_k \]

because the denominator is identical for all classes.

LDA assumes that the density is Gaussian:

\[f_k(x) = \frac{1}{(2 \pi)^{\frac{p}{2}} \sqrt{|\Sigma_k|}} \exp\left(-\frac{1}{2}(x - \mu_k)^T\Sigma_k^{-1}(x - \mu_k)\right)\]

where $\Sigma_k$ is the covariance matrix for the samples from class $k$ and $|\Sigma_k|$ is its determinant. LDA assumes that all classes have the same covariance matrices, i.e. $\Sigma_k = \Sigma\,, \forall k$. 

Plugging $f_k$ into the classification function, we find that

\[G(x) = \arg \max_k \delta_k(x)\] 

where 

\[\delta_k(x) = x^T \Sigma^{-1} \hat{\mu}_k -\frac{1}{2} \hat{\mu}_k^T \Sigma^{-1} \mu_k + \log \pi_k \]

is the discriminant function for class $k$. So, now that we have a classifier, how we can compute it?

To find the covariance matrix, we simply compute

\[\Sigma = \sum_{k=1}^K \frac{1}{N - K} \sum_{g_i = k} (x_i - \hat{\mu}_k) (x_i - \hat{\mu}_k)^T\,.\]

Note that the deviation from the means is divided by $N-K$, the degrees of freedom, to obtain an unbiased estimator.

The means of the classes, which are also called centrois, are defined by 

\[\hat{\mu}_k = \frac{1}{N_k} \sum_{g_i = k} x_i\,.\]

The priors are set to the ratio of the class-specific observations:

\[\hat{\pi}_k = \frac{N_k}{N}\,.\]

With this, we have defined all parameters required for the classifier. We have, however, not yet dealt with the dimensionality reduction that LDA performs. This procedure involves the within-class variance, $W$, and the between-class variance, $B$. The between-class variance indicates the deviation of centroids from the overall mean, $\hat{\mu} = \sum_{k=1}^K \hat{\pi}_k \hat{\mu}_k$, and is defined as:

\[B = \sum_{k=1}^K \hat{\pi}_k(\hat{\mu}_k - \hat{\mu}) (\hat{\mu}_k - \hat{\mu})^T\,.\]

Finding a sequence of optimal substeps involves three steps:


1. Compute the $K \times p$ matrix $M$ containing the centroids, $\mu_k$, and determine the common covariance matrix $W = \Sigma$.
2. Compute $M^{\ast} = MW^{-\frac{1}{2}}$ using the eigen-decomposition of W.
3. Compute $B^{\ast}$ (the between-class covariance), the covariance matrix of $M^{\ast}$, and its eigen-decomposition $B^{\ast} = V^{\ast} D_B {V^{\ast}}^T$. The columns $v^{\ast}_l$ of $V^{\ast}$ define the coordinates of the reduced subspace.

Thus, the $l$-th discriminant variable (one of the $K-1$ new dimensions) is determined by $Z_l = v_l^T X$ with $v_l = W^{-\frac{1}{2}} v^{\ast}_l$.

Next, we will deal with the alternative formulation of LDA due to Fisher, which provides a better intuition on the canonical variables.

### Interpretation as a transformation

Fisher's LDA optimization criterion states that the centroids of the groups should be spread out as far as possible. This amounts to finding a linear combination $Z = a^T X$ such that $a^T$ maximizes the between-class variance relative to the within-class variance.

The within-class variance is $W$ is simply the overall covariance matrix, $\hat{\Sigma}$. This matrix determines the deviation of all observations from their class centroids. The between-class variance is defined according to the deviation of the centroids from the overall mean, as defined earlier.
For $Z$, the between class variance is $a^T B a$ and the within-class variance is $A^T W a$. Thus, LDA can be optimized through the Rayleigh quotient

\[\max_a \frac{a^T B a}{a^T W a}\,, \]

which defines an optimal mapping of $X$ to the new space $Z$. Note that $Z \in \mathbb{R}^{1 \times p}$, that is, the observations are mapped to a single dimension. To obtain additional dimensions, we need to solve the optimization problem for $a_1, \ldots, a_{K-1}$ where each successive $a_k$ is constructed to be orthogonal in $W$ to the previous discriminant coordinates. This leads to the linear transformation $G = (Z_1^T, Z_2^T, \ldots, Z_{K-1}^T) \in \mathbb{R}^{p \times q}$ with which we can map $X$ from $p$ to $q$ dimension via $X G$. Why do we consider $K-1$ projections? This is because the affine subspace that is spanned by the $K$ centroids has a rank of at most $K-1$ due to the spherical assumptions of LDA.

The benefit of this formulation of LDA is that is enables us to perform classification in a reduced subspace. We do not need to use the full $K-1$ dimensions but can choose a smaller subspace $l < K-1$. This is called reduced-rank LDA. The motivation for reduced-rank LDA is that classification basd on a reduced number of discriminant variables can improve performance on the test set when the model is overfitted to the training data.

Using Fisher's formulation of LDA, classification involves two steps:

1. Sphere the data using the common covariance matrix $\hat{\Sigma} = UDU^T$ (eigendecomposition) such that $X^{\ast} = D^{-\frac{1}{2}} U^T X$. In this way, the covariance of $X^{\ast}$ becomes the identity matrix. By eliminating the covariance between the variables through sphering the data, it becomes much easier for separation of classes in the transformed space.
2. Classify observations $x_i$ to the closest class centroid in the transformed space, taking into account the class priors $\pi_k$. The intuition about the $\pi_k$ is that if an observation would have equal distance to the centroids from two classes, then it would be assigned to the class with the greater prior.

## Summary

In summary, LDA can be understood in two complementary ways. The first view focuses on the Gaussian assumption, while the second view focuses on the transformation to canonical variables.

### Probabilistic view

LDA uses Bayes' rule to determine the posterior probability that an observation $x$ belongs to class $k$. Due to the normal assumption of LDA, the posterior is defined through Gaussian densities where the covariance matrix is assumed to be identical for all of the classes. New points are classified by computing the discriminant function $\delta_k$ (the enumerator of the posterior probability) for each class and returning the class $k$ with maximal $\delta_k$. The discriminant variables
can be obtained through eigen-decompositions of the within-class and between-class variance.

### Fisher's view

According to Fisher, LDA can be understood as a dimensionality reduction where each successive transformation is orthogonal and maximizes the between-class variance relative to the within-class variance. This transforms the feature space to an affine space with $K-1$ dimensions. After sphering the input data, new points can be classified by determining the closest centroid in the affine space under consideration of the class priors.


## Complexity of the LDA model

The number of effective parameters of LDA can be derived in the following way.
There are $K$ means, $\hat{\mu}_k$ that are estimated. The covariance matrix does not require additional parameters because it is already defined by the centroids. Since we need to estimate $K$ discriminant functions (to obtain the decision boundaries), this gives rise to $K$ calculations involving the $p$ elements. Additionally, we have $K-1$ free parameters for the $K$ priors. Thus, the number of effective LDA parameters is $Kp + (K-1)$. 

## Properties of LDA

LDA has the following properties:

* LDA assumes that the data are Gaussian. More specificially, it assumes that all classes share the same covariance matrix. 
* LDA is a linear method that is not suited if there are higher-order interactions between the independent variables.
* LDA classifies samples by mapping them to $K-1$ discriminant variables. Since observations are represented in terms of few discriminant variables, LDA is well-suited for multi-class problems. 
* LDA is unsuited when the class distribution is imbalanced since the priors are estimated from the observed counts. Thus, observations will rarely be classified to infrequent classes.
* Similarly to PCA, LDA can be used as a dimensionality reduction technique. Note that the transformation of LDA is inherently different to PCA because LDA is a supervised method that considers the outcomes.

## The phoneme data set

We will use the [phoneme speech recognition data set](https://web.stanford.edu/~hastie/ElemStatLearn/datasets/phoneme.info.txt):

```{r, message = FALSE}
library(RCurl)
f <- getURL('https://www.datascienceblog.net/data-sets/phoneme.csv')
df <- read.csv(textConnection(f), header=T)
print(dim(df))
```

The data set contains samples of digitized speech for five phonemes: aa (as the vowel in *dark*), *ao* (as the first vowel in *water*), *dcl* (as in *dark*), *iy* (as the vowel in *she*), and *sh* (as in *she*). In total, 4509 speech frames of 32 msec were selected. For each speech frame, a log-periodogram of length 256 was computed, which is suitable for speech recognition. The 256 columns labeled *x.1* to *x.256* identify the speech features, while the columns *g* and *speaker* indicate the phonemes (labels) and speakers, respectively.

Let us assign training and test samples:
```{r}
#logical vector: TRUE if entry belongs to train set, FALSE else
train <- grepl("^train", df$speaker)
# remove non-feature columns
to.exclude <- c("row.names", "speaker", "g")
feature.df <- df[, !colnames(df) %in% to.exclude]
test.set <- subset(feature.df, !train)
train.set <- subset(feature.df, train)
train.responses <- subset(df, train)$g
test.responses <- subset(df, !train)$g
```

## Fitting an LDA model in R

We can fit an LDA model in the following way:

```{r}
library(MASS)
lda.model <- lda(train.set, grouping = train.responses)
```

We can transform the training data to the canonical coordinates in the following way:

```{r}
lda.prediction.train <- predict(lda.model, train.set)
x.projected <- lda.prediction.train$x # feature transformation 
# visualize the features in the two LDA dimensions
plot.df <- data.frame(x.projected[,1:2], "Outcome" = train.responses)
library(ggplot2)
ggplot(plot.df, aes(x = LD1, y = LD2, color = Outcome)) + geom_point()
```

Plotting the data in the two LDA dimensions reveals three clusters:

* Cluster 1 (left) consists of *aa* and *ao* phonemes
* Cluster 2 (bottom right) consists of *dcl* and *iy* phonemes
* Cluster 3 (top right) consists of *sh* phonemes

This indicates that two dimensions are not sufficient for differentiating all 5 classes. However, the clustering indicates that phonemes that are sufficiently different from one another can be differentiated very well. 

We can also plot the mapping of training data onto the discriminant dimensions using the ```plot.lda``` function where the *dimen* parameter can be used to specify the number of considered dimensions:

```{r}
colors <- c("red", "blue", "green", "yellow", "black") # aa, ao, dcl, iy, sh
my.cols <- colors[match(lda.prediction.train$class, levels(df$g))]
plot(lda.model, dimen = 4, col = my.cols)
```

Plotting the training data for all dimension pairs demonstrates that, by construction, the first two dimensions separate the phoneme groups best. What we can learn from the plot, though, is how many dimensions we should select for a reduced-rank LDA. Remember that LD1 and LD2 confused *aa* with *ao* and *dcl* with *iy*. Thus, we would like additional dimensions that help us differentiating these groups. It seems that we need all of the four dimensions because *dcl* and *iy* are only well-separed in LD1 vs LD3, while *aa* and *ao* are only well-separated when LD4 is combined with any of the other dimensions.

To visualize the centroids of the groups, we can create a custom plot:

```{r, fig.height = 8, fig.width = 12}
plot_lda_centroids <- function(lda.model, train.set, response) {
    centroids <- lda.model$means %*% lda.model$scaling 
    library(RColorBrewer)
    colors <- brewer.pal(8, "Accent")
    my.cols <- colors[match(lda.prediction.train$class, levels(df$g))]
    my.points <- as.matrix(train.set) %*% lda.model$scaling
    no.classes <- length(levels(response))
    par(mfrow=c(no.classes -1, no.classes -1), mar=c(1,1,1,1), oma=c(1,1,1,10))
    for (i in 1:(no.classes - 1)) {
        for (j in 1:(no.classes - 1)) {
            y <- my.points[, i]
            x <- my.points[, j]
            cen <- cbind(centroids[, j], centroids[, i])
            if (i == j) {
                plot(x, y, type="n") 
                max.y <- max(my.points[, i])
                max.x <- max(my.points[, j])
                min.y <- min(my.points[, i])
                min.x <- min(my.points[, j])
                max.both <- max(c(max.x, max.y))
                min.both <- max(c(min.x, min.y))
                center <- min.both + ((max.both - min.both) / 2)
                text(center, center, colnames(my.points)[i], cex = 3)}
            else {
                plot(x, y, col = my.cols, pch = as.character(response), xlab ="", ylab="")
                points(cen[,1], cen[,2], pch = 21, col = "black", bg = colors, cex = 3)
            }
        }
    }
    par(xpd = NA)
    legend(x=par("usr")[2] + 1, y = mean(par("usr")[3:4]) + 20, 
            legend = rownames(centroids), col = colors, pch = rep(20, length(colors)), cex = 3)
}
plot_lda_centroids(lda.model, train.set, train.responses)
```

## Reduced-rank LDA

Although this data set is not optimal for reduced-rank LDA since it seems that all dimensions are worthwhile, we will still showcase the use of reduced-rank LDA by evaluating the accuracy of LDA based on one to four discriminant coordinates:

```{r}
# perform LDA with 1 to 4 discriminant coordinates
dims <- 1:4
accuracies <- rep(NA, length(dims))
for (i in seq_along(dims)) {
    lda.pred <- predict(lda.model, test.set, dim = dims[i])
    acc <- length(which(lda.pred$class == test.responses))/length(test.responses)
    accuracies[i] <- acc
}
print(accuracies)
```

As expected from the visual exploration of the transformed space, the test accuracy increases with each additional dimension. Since LDA with four dimension obtains the maximal accuracy, we would decide to use all of the discriminant coordinates for classification. 

We can visualize the performance of the classifier using four dimensions:

```{r}
lda.pred <- predict(lda.model, test.set)
plot.df <- data.frame(lda.pred$x[,1:2], "Outcome" = test.responses, 
                    "Prediction" = lda.pred$class)
ggplot(plot.df, aes(x = LD1, y = LD2, color = Outcome, shape = Prediction)) +
        geom_point()
```

In the plot, the labels are indicated by different colors and the predictions by different colors. Ideally, each color would have only a single symbol. Incorrect predictions are visible when one colors exhibits different symbols. Using the plot, we quickly see that most confusions occur when observations labeled as *aa* are incorrectly classified as *ao* and vice versa.

## Quadratic discriminant analysis

Since we have already covered LDA in detail, it is very easy to explain QDA. QDA allows the covariance matrices to be different for each class. This means we need to estimate $\Sigma_k$ for each class $k \in \{1, \ldots, K\}$ rather than assuming $\Sigma_k = \Sigma$ as in LDA. The discriminant function of LDA is quadratic in $x$:

\[\delta_k(x) = - \frac{1}{2} \log |\Sigma_k| - \frac{1}{2}(x- \mu_k)^ \Sigma_k^{-1} (x - \mu_k) + \log \pi_k\]

Since QDA estimates a covariance matrix for each class, it has a greater number of effective parameters than LDA. We can derive the number of parameters in the following way.

* We need $K$ class priors $\pi_k$. Since $\sum_{i=1}^K \pi_k = 1$, we do not need a parameter for one of the priors. Thus, there are $K-1$ free parameters for the priors.
* The $K$ centroids $\mu_k$ with $p$ entries each give $Kp$ parameters.
* A covariance matrix $\Sigma_k$. From the matrix, we only need to consider the diagonal and the upper right triangle. This region of the covariance matrix has $\frac{p (p+1)}{2}$ elements. Since $K$ such matrices need to be estimated, these are $K \frac{p (p+1)}{2}$ parameters.

So, the effective number of QDA parameters is $K-1 + Kp + K \frac{p (p+1)}{2}$. Since the number of QDA parameters is quadratic in $p$, QDA should be used with care when the feature space is large. QDA is particularly useful if there is prior knowledge that individual classes exhibit distinct covariances. 

## Regularized discriminant analysis

RDA is a compromise between LDA and QDA as it shrinks $\Sigma_k$ to a pooled variance $\Sigma$ by defining

\[\hat{\Sigma}_k(\alpha) = \alpha \hat{\Sigma}_k + (1 - \alpha) \hat{\Sigma}\]

and replacing $\hat{\Sigma}_k$ with $\hat{\Sigma}_k(\alpha)$ in the discriminant functions. Here, $\alpha \in [0,1]$ is a tuning parameter determining whether the covariances should be estimated independently ($\alpha = 1$) or should be pooled ($\alpha = 0$).

## Refs
https://onlinecourses.science.psu.edu/stat857/node/75/
https://web.stanford.edu/class/stats202/content/lec9.pdf
https://machinelearningmastery.com/linear-discriminant-analysis-for-machine-learning/


