---
title: "Improving the whiskey distillery data set"
author: Matthias Döring
date: '2018-11-13T18:00:00Z'
description: "In this post, I clean up and augment a data set that provides their taste characteristics. The improved data set is augmented with the regions where the distilleries are situated, as well as their geological location in terms of longitude and latitude."
categories:
  - Other

---

I have previously used a data set describing the characteristics of whiskeys to [draw radar plots](/post/data-visualization/radar-plot/). Here, I present how I cleaned and augmented [the original data from the University of Strathclyde](http://outreach.mathstat.strath.ac.uk/outreach/nessie/datasets/whiskies.txt), resulting in an [improved version of the whiskey data set](https://www.datascienceblog.net/data-sets/whiskies.csv).

## Loading the whiskey data set

The original data set can be loaded from the web in the following way:

```{r, message = FALSE}
library(RCurl)
# load data as character
f <- getURL('https://www.datascienceblog.net/data-sets/whiskies.txt')
# read table from text connection
df <- read.csv(textConnection(f), header=T)
```

## Fixing the post codes

Since there are tab characters and spaces in the post codes, we will clean those up:

```{r}
head(df)
df$Postcode <- gsub(" *\t*", "", df$Postcode)
```

## Annotating the locations of the distilleries

Now, we will retrieve the locations of the distilleries:

```{r, message = FALSE, fig.width = 10, fig.height = 10}
# transform UTM coordinates to longitude/latitude in degrees
geo.df <- df[, c("Latitude", "Longitude")]
colnames(geo.df) <- c("lat", "long") # switch for plotting
# REF: https://rstudio-pubs-static.s3.amazonaws.com/375481_d2605e8726364d30a5b20c6974710cf5.html
# create 'SpatialPoints' object
library(raster)
coordinates(geo.df) <-  ~lat + long
# add coordinate reference system (CRS) for UK
proj4string(geo.df) <- CRS("+init=epsg:27700")
# transform again:
# NB: getting rgdal working on old systems is tough due to libgdal dependency
library(rgdal)
geo.df <- spTransform(geo.df, CRS("+init=epsg:4326"))
map.df <- data.frame("Distillery" = df[, "Distillery"], geo.df)
df <- cbind(df, map.df[, c("lat", "long")])
```

## Annotation of regions

To annotate the regions in which the distilleries are situaded, I relied on a [list of Scottisch distilleries available at Wikipedia](https://en.wikipedia.org/wiki/List_of_whisky_distilleries_in_Scotland).
