---
title: "Determining the Distribution of Data Using Histograms"
author: Matthias Döring
date: '2018-10-28'
description: "Histograms are an ideal tool for visualizing the distribution of a variable and frequently used for data exploration. Here, I show how a histogram acan aid in differentiating two distributions."
categories:
  - data-visualization
tags:
  - plot
thumbnail: "post/data-visualization/histogram_cover.png"

---
It is always useful to spend some time exploring a new data set before processing it further and analyzing it. One of the most convenient ways to get a feel for the data is plotting a histogram. The histogram is a tool for visualizing the frequency of measurements in terms of a bar plot. Here we'll take a closer look at how the histogram can be used in R.
<!--more-->

## Plotting a histogram in native R

The basic histogram fun is `hist`. It is very easy to use because you only need to provide a numeric vector. For example:

```{r}
# fix seed for reproducibility
set.seed(1)
# draw n samples from the uniform distribution in the range [0,1]
n <- 100
x <- runif(n)
hist(x)
```

If you need a higher resolution, you can increase the number of bins using the `breaks` argument:

```{r}
hist(x, breaks = 20)
```

These results indicate that although the samples were drawn from the uniform distribution, there are still some values that are over- and underrepresented.

## Plotting a histogram using ggplot

If you want to have more control over your plots, then you should use the ggplot2 library, which is part of the [tidyverse suite](https://www.tidyverse.org/). Since the plotting function expects a data frame, we'll have to construct one first:

```{r, message = FALSE}
library(ggplot2)
df <- data.frame(x = x)
# map data frame values in column x to x-axis and plot
ggplot(df, aes(x = x)) + geom_histogram()
```

### Plotting a histogram for several groups

To exemplify why ggplot is great, let's assume we have more than one vector we'd like to plot. This can be done in the following way:

```{r, message = FALSE}
# normally distributed data
y <- rnorm(n)
df <- data.frame(Category = c(rep("Uniform", n), rep("Normal", n)), 
                x = c(x, y))
# fill bars according to 'Category'
ggplot(df, aes(x = x, fill = Category)) + geom_histogram()
```

As an alternative to plotting the data within a single plot, one can also use a faceted plot:

```{r, message = FALSE}
ggplot(df, aes(x = x)) + geom_histogram() + facet_wrap(vars(Category))
```

One problem with the facet problem is that the scales may be different for the two axes. For example, the data from the normal distribution have a smaller maximum frequency than those from the uniform distribution. Thus, values from the normal distribution are not clearly visible. Moreover, the uniform data are in a smaller value range than the normal data. To fix this problem, we can provide an additional argument that makes both the x-axis and the y-axis flexible:

```{r, message = FALSE}
ggplot(df, aes(x = x)) + geom_histogram() + 
        facet_wrap(vars(Category), scales = "free")
```

Now, the difference between the two distributions is much more easily visible than before.
